Investigating Overfitting in Convolutional Neural Networks

### 2025-05-27 paper_formatted.tex
push to fix brier's version

%! Author = JustinMa
%! Date = 5/22/25

@@

    \section*{1. Introduction}

    Neural networks are a form of machine learning models, roughly inspired by how the human brain
    processes information. They have become increasingly prevalent in our developing world, providing
    powerful and unique solutions to a wide variety of problems. However, with power also comes risk,
    as training a neural network too much on a small dataset can cause overfitting. Overfitting occurs
    when a model learns the training data too well by essentially memorizing it, causing it to perform
    well on data in its training set but poorly on test data (data that were not in the training set).
    While it might seem that simply increasing a model’s size would always boost accuracy by allowing it to
    capture more patterns, extra parameters often give the network enough capacity to store the training set
    outright when the training set is small, allowing it to overfit. Smaller networks, by contrast, are forced to find the underlying patterns
    in the data because of their limited size. In this study, we investigate whether increasing the number of
    parameters in a convolutional neural network leads to greater overfitting when trained on a small dataset, as
    measured by the difference between its accuracy on the training dataset and the test dataset.

    \section*{2. Statistical Question}

@@

    \section*{3. Data Collection}

@@

    \noindent Each model was trained for 100 epochs on the randomly selected subset of 100 images, then evaluated
    on the full test set of the CIFAR-10 dataset, made up of 10,000 images. The difference between the train and test accuracy (train - test) was then
    computed for each model and graphed on a scatter plot with the x-axis
    as parameter count and the y-axis as the difference between train and test accuracy for each model.

    \section*{4. Data Display}
    \begin{figure}[H]
        \centering
        \includegraphics[width=1\textwidth]{Images/Scatter}
        \caption{Scatterplot of Parameter Count vs. Difference in Train and Test Accuracy (Train - Test)}
        \label{fig:scatterplot}
    \end{figure}

@@

    \noindent
    Based on the scatter plot (\textit{\autoref{fig:scatterplot}}) and the correlation coefficient of $r=\sqrt{0.1037}=0.3220$ (\textit{from \autoref{tab:minitab}}), there appears to be a weak,
    positive, linear relationship between parameter count and difference between train and test accuracy (train - test).
    There appear to be a few possible high outliers above $x =$ 1.5 million parameters
    and a few possible low outliers above $x = 0.5$ million parameters.
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.7\textwidth]{Images/Resid}
        \caption{Residual Plot (Parameter Count vs. Residuals)}
        \label{fig:residuals}
    \end{figure}
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.7\textwidth]{Images/Resid Hist}
        \caption{Histogram of Residuals}
        \label{fig:residual_hist}
    \end{figure}
    \noindent The histogram of the residuals appears to be approximately symmetric, with a median around 0 and a range of
    $0.05-(-0.04)=0.09$. There does not appear to be any strong skew or outliers.

    \section*{5. Data Analysis}

    We conducted a $t$-test for the population slope $\beta$ of the least-squares regression line that relates the number of parameters in a model trained on a 100-image subset of the CIFAR-10 dataset to the difference in accuracy between training and test data (Train Accuracy $-$ Test Accuracy).

@@

    \subsubsection*{5.2.2 $t$-statistic}
    \noindent And now we can find the t-statistic, which represents how many standard deviations away our sample slope $b$ is from the population slope $\beta$ on the sampling distribution, assuming that $\beta=0$:
    \begin{gather*}
        \mathrm{t} = \frac{b}{SE_\beta} \\[1em]
        \mathrm{t} = \frac{0.008312}{0.002008} = 4.139
    \end{gather*}

    \subsubsection*{5.2.3 $p$-value}
    \noindent Now that we have the t-statistic, we can use it to find the $p$-value, which represents the probability of getting a test statistic
    of $4.139$ or higher, assuming that the population slope $\beta$ is $0$. Because we have 150 data points, we have $150 - 2 = 148$ degrees of freedom.
    Thus, we calculate the $p$-value by finding the amount of area under a $t$-distribution with 148 degrees of freedom, above $t=4.139$. The graph below demonstrates this:

@@

    \noindent Based on the area under the curve in \textit{\autoref{fig:t_dist}} above, $t=4.139$, our $p$-value is $2.916\cdot10^{-5}$.

    \section*{6. Conclusion}

    The $p$-value of $2.916\cdot10^{-5}$ indicates that, assuming that the population slope $\beta=0$, there is around a
    $0.002916\%$ probability of getting a sample slope as great or greater than our sample slope of $0.008312$, in a sample of
    150 randomly selected models from the population of convolutional models of our structure and of size 1 million parameters to 25 million parameters, trained
    on a randomly selected subset of 100 images from the CIFAR-10 train dataset and tested on the full test dataset of CIFAR-10.
    Because the $p$-value of $2.916\cdot10^{-5}$ is less than $\alpha=0.05$, we reject the null hypothesis, which states that the population slope $\beta$
    of the least-squares regression line that relates parameter count to difference in train and test accuracy (train - test) is $0$,
    in the population defined above.
    The data do provide convincing evidence that the population slope $\beta$ is greater than $0$, and that there is a positive correlation between
    parameter count and difference between train and test accuracy (train - test) in the population defined above. Thus, we can conclude that
    increasing the number of parameters in the model does indeed cause overfitting.

    \section*{7. Reflection}

>>>>

%! Author = JustinMa
%! Date = 5/27/25

@@

    \section*{1 Introduction}

    Neural networks are a form of machine learning models, inspired by how the human brain
    processes information. They have become increasingly prevalent in our developing world, providing
    powerful and unique solutions to a wide variety of problems. However, with power also comes risk,
    as training a neural network too much on a small dataset can cause overfitting. Overfitting occurs
    when a model learns the training data too well by essentially memorizing it, causing it to perform
    well on data in its training set but poorly on test data (data that were not in the training set).
    While it might seem that simply increasing a model’s size would always boost accuracy by allowing it to
    capture more patterns, extra parameters often give the network enough capacity to store the training set
    outright when the training set is small, allowing it to overfit. Smaller networks, by contrast, are forced to find the underlying patterns
    in the data because of their limited size. In this study, we investigate whether increasing the number of
    parameters in a convolutional neural network leads to greater overfitting when trained on a small dataset, as
    measured by the difference between its accuracy on the training dataset and the test dataset.

    \section*{2 Statistical Question}

@@

    \section*{3 Data Collection}

@@

    \noindent Each model was trained for 100 epochs on the randomly selected subset of 100 images, then evaluated
    on the full test set of the CIFAR-10 dataset, made up of 10,000 images. The difference between the train and test accuracy (train - test) was then
    computed for each model and graphed on a scatter plot with the $x$-axis
    as parameter count and the $y$-axis as the difference between train and test accuracy for each model.

    \section*{4 Data Display}
    \begin{figure}[H]
        \centering
        \includegraphics[width=1\textwidth]{Images/Scatter}
        \caption{Scatterplot of Parameter Count vs. Difference in Train and Test Accuracy (Train - Test)}
        \label{fig:scatterplot}
    \end{figure}

@@

    \noindent
    Based on the scatter plot (\textit{\autoref{fig:scatterplot}}) and the correlation coefficient of $r=\sqrt{0.1037}=0.3220$ (\textit{from \autoref{tab:minitab}}), there appears to be a weak,
    positive, linear relationship between parameter count and difference between train and test accuracy (train - test).
    There appear to be a few possible high outliers above $x =$ 1.5 million parameters
    and a few possible low outliers above $x = 0.5$ million parameters.
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.6\textwidth]{Images/Resid}
        \caption{Residual Plot (Parameter Count vs. Residuals)}
        \label{fig:residuals}
    \end{figure}
    \noindent Because the residual plot (\textit{\autoref{fig:residuals}}) appears to have no form and because the standard
    deviation of the residuals $S$ (\textit{from \autoref{tab:minitab}}) is small ($0.01703$), the graphical display seems to
    indicate that the null hypothesis will be rejected.
    \begin{figure}[H]
        \centering
        \includegraphics[width=0.6\textwidth]{Images/Resid Hist}
        \caption{Histogram of Residuals}
        \label{fig:residual_hist}
    \end{figure}
    \noindent The histogram of the residuals appears to be approximately symmetric, with a median around 0 and a range of
    $0.05-(-0.04)=0.09$. There does not appear to be any strong skew or outliers.

    \section*{5 Data Analysis}

    We conducted a $t$-test for the population slope $\beta$ of the least-squares regression line that relates the number of parameters in a model trained on a randomly selected 100-image subset of the CIFAR-10 dataset to the difference between its accuracy on training and test data (Train Accuracy $-$ Test Accuracy).

@@

    \subsubsection*{5.2.2 $t$-statistic}
    \noindent And now we can find the $t$-statistic, which represents how many standard deviations away our sample slope $b$ is from the population slope $\beta$ on the sampling distribution, assuming that $\beta=0$:
    \begin{gather*}
        \mathrm{t} = \frac{b}{SE_\beta} \\[1em]
        \mathrm{t} = \frac{0.008312}{0.002008} = 4.139
    \end{gather*}

    \subsubsection*{5.2.3 $p$-value}
    \noindent Now that we have the $t$-statistic, we can use it to find the $p$-value, which represents the probability of getting a test statistic
    of $4.139$ or higher, assuming that the population slope $\beta$ is $0$. Because we have 150 data points, we have $150 - 2 = 148$ degrees of freedom.
    Thus, we calculate the $p$-value by finding the amount of area under a $t$-distribution with 148 degrees of freedom, above $t=4.139$. The graph below demonstrates this:

@@

    \noindent Based on the area under the curve in \textit{\autoref{fig:t_dist}} above $t=4.139$, our $p$-value is $2.916\cdot10^{-5}$.

    \section*{6 Conclusion}

    The $p$-value of $2.916\cdot10^{-5}$ indicates that, assuming that the population slope $\beta=0$, there is around a
    $0.002916\%$ probability of getting a sample slope as great or greater than our sample slope of $0.008312$, in a sample of
    150 randomly selected models from the population of convolutional models of our structure and of size 1 million parameters to 25 million parameters, trained
    on a randomly selected subset of 100 images from the CIFAR-10 train dataset and tested on the full test dataset of CIFAR-10.
    Because the $p$-value of $2.916\cdot10^{-5}$ is less than $\alpha=0.05$, we reject the null hypothesis, which states that the population slope $\beta$
    of the least-squares regression line that relates parameter count to difference in train and test accuracy (train - test) is $0$,
    in the population defined above.
    The data do provide convincing evidence that the population slope $\beta$ is greater than $0$, and that there is a positive correlation between
    parameter count and difference between train and test accuracy (train - test) in the population defined above. Thus, for that population, we can conclude that
    increasing the number of parameters in the model does indeed cause overfitting.

    \section*{7 Reflection}
